---
title: "Data Analysis Project"
author: "Group 3"
date: "2025-11-22"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

In this notebook, we are coding Maryland Notices of Intent to Foreclose by Zip Code from Maryland's Open Data Portal. Link to URL: <https://opendata.maryland.gov/Housing/Maryland-Notices-of-Intent-to-Foreclose-by-Zip-Cod/ftsr-vapt/about_data> 

## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}
options(scipen=999) 

library(readr)
library(tidyverse)
library(lubridate)
library(janitor)
library(tidycensus)
```

## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary for data you are working with, and discuss any caveats or issues you discovered working with this data. 

About: The Maryland Notices of Intent to Forclose by Zip Code dataset contains 53 columns and 563 rows of data (updates periodically). Each row represents a single Maryland zip code, and the columns correspond to months from July 2021 through October 2025, containing the number of foreclosure notices reported for each zip code during that month. The first column is 'zip,' which identifies the zip code. The remaining 52 columns correspond to monthly foreclosure counts. The dataset was originally created on July 12, 2022. 

The dataset is compiled from a third-party report from the Maryland Office of Financial Regulation, and may contain duplicates or errors. The financial regulation office emphasized that the data may differ from official foreclosure court records, so totals may not perfectly reflect actual foreclosure activity. 

The data updates, but what we have here covers July 2021 to October 2025, roughly four years. Any analysis backed with data would not be ideal outside of this timeframe. 

The dataset does not include demographic, economic or housing characteristics by zip code (like population, income or racial group). To answer questions about income or racial demographics, we will need to join U.S. Census data with the foreclosure dataset. No information about property type, loan type or reasons for foreclosure is included in the original dataframe. Additionally, the dataset provides counts, not rates relative to population or housing units. Again, we'll merge data from ACS to analyze this data properly. 


Variables: 
- 'zip': Every zipcode across Maryland. 
- When a column has a month followed by an underscore and then the year (ex.: 'july_2021'), it means, in that column, it will include the total number of foreclosure notices that occurred during that specific month and year for each zip code. 

Notes and Caveats: 
- It's important to note that these are notices of intent to foreclose, even though we may shorten that in code as foreclosure notices or 'foreclose_notice.' These foreclosure notice counts do not equal to foreclosure case records. 
- Some zip codes have zero foreclosure notices for a month(s), but others may have several. 
- There is no total amount because the dataset is frequently updated. 
- Any duplicates or errors may be present. 
- Last update was Nov. 6, 2025. 
- Exported this data on Nov. 22, 2025, when the last update still read Nov. 6, 2025. 
- At a glance, the dataset allows us to:
  - Calculate the total foreclosures by zipcode. 
  - Identify trends across months and years.
  - Compare foreclosure notices between zipcodes.
  - Create choropleth maps showing foreclosure totals or per capita rates by zip. 
- The dataset simply holds the totals. It does not include any specific information about the notices of intent to foreclose besides the zip code, month and year. 
- We will need to join census data for demographic/rate questions. Everything else should be in this dataset.
  - 'zip' is correctly listed as a character, so this way we can join it easily. If it is needed as a number, then we must change it. 
  - All the months are listed as a numeric data type. 
  - We need to make more aggregated dataframes for our analyses.
  
  
  

```{r}
# Load required data
# Path to data should be loaded from folder "data" i.e. read_csv("data/name_of_data.csv")
# Clean required data and prepare for analysis if needed. 

md_foreclosure_notices_zipcode <- read_csv("data/md_foreclosure_notices_zipcode.csv") |>
  clean_names()

md_foreclosure_notices_zipcode

glimpse(md_foreclosure_notices_zipcode)
```

## Sentences to Code
In this notebook, we are coding five sentences using the state data.

### Sentence 1

* **Which zipcode has the highest number of foreclosure notices total, and which has the highest rate over the four year period? Where are those areas?**
* **XXXX**: The zip code 20772 had the most notices to foreclose of the dataset at 6,877 and therefore the highest rate of 132.25 notices per month. It's located in Prince George's County and encompasses much of Upper Marlboro. The next highest is the area neighboring directly North, zip code 20774 at 5,856 notices of 112.62 per month. The zipcode with the third most notices and the last with a rate above 100 is 21206 in Baltimore City, with 5551 notices at 106.75 a month. Generally the top zipcodes appear to be in Prince Goerge's or Baltimore City and County. 

```{r}
# Put code here
md_foreclosures_long <- md_foreclosure_notices_zipcode |>
  pivot_longer(cols = contains("_"), names_to = "month", values_to = "foreclosures") |>
  mutate(
    month = dmy(paste("01", str_replace(month, "_", " "))),
    year = year(month)
  )

# Use above dataframe for analysis but don't modify - date columns for month and year made and pivoted for easier use
  
md_foreclosures_summary <- md_foreclosures_long |>
  group_by(zip) |>
  summarise(total = sum(foreclosures, na.rm = TRUE),
            months = n(),
            avg_per_month = total / months
  ) |>
  arrange(desc(total))
# Display results of code below this codeblock
md_foreclosures_summary
```

### Sentence 2

* **What month has the highest number of foreclosure notices? What about year? Are there any differences or similarities between the month that has the highest and the year that has the highest? Why might this be?**
* **XXXX**: 2021 is dwarfed by the other years at 14027 foreclosure notices. This could be because it is only a half year of data (July - December), but it still seems small compared to the next highest year, 2022, at 55369. The notices seem to increase every year, with the only discrepancy being 2025 under 2024 with 74930 versus 79577 notices. However, 2025 is also an incomplete year with its data ending in October, so the trend seems legitimate. The top three months are in 2025, and generally the months themselves don't seem to correlate much with any trends. 

```{r}
# Put code here
month_foreclosures <- md_foreclosures_long |>
  group_by(month) |>
  summarise(total = sum(foreclosures, na.rm = TRUE),
  ) |>
  arrange(desc(total))

year_foreclosures <- md_foreclosures_long |>
  group_by(year) |>
  summarise(total = sum(foreclosures, na.rm = TRUE),
  ) |>
  arrange(desc(total))

# Display results of code below this codeblock
month_foreclosures 
year_foreclosures
```

### Sentence 3

* **What are the majority racial demographics of zip codes with notably high or low foreclosure rates? How about their median income?**
* **XXXX**: 9 of the top 10 zip codes are majority black, and the case seems the same for the majority of zip codes with above 50 foreclosure notices per month. The bottom of the list has many zip codes with unavailable demographic statistics but as they start to show up, white majority zips become frequent. The top two zip codes have very high median incomes but beyond that there doesn't seem to be any correlation to notices as the statistic begins to vary widely. 

```{r}
#Make sure API is installed: census_api_key("your_key", install = TRUE)
#Put code here
race_income <- get_acs(
  geography = "zcta",
  variables = c(
    median_income = "B19013_001",
    total_pop = "B02001_001",
    white = "B02001_002",
    black = "B02001_003",
    native = "B02001_004",
    asian = "B02001_005"
  ),
  year = 2022,
  survey = "acs5",
  output = "wide"
) |>
  rename(zip = GEOID) |>
  mutate() |>
  mutate(
    zip = as.character(zip),
    pct_white = whiteE / total_popE,
    pct_black = blackE / total_popE,
    pct_asian = asianE / total_popE,
    pct_native = nativeE / total_popE
    )
  
census_foreclosures_summary <- md_foreclosures_summary |>
  left_join(race_income, by = "zip") |>
  arrange(desc(total))

# Display results of code below this codeblock
census_foreclosures_summary

```

### Sentence 4

* **Which zip codes have had the largest increase in foreclosures from July 2021 to October 2025? Which had the largest decrease? Are these in similar areas?**
* **XXXX**: 20772 had the largest increase in foreclosure notices from July 2021 to October 2025, with a 199 foreclosure difference. While, 20618 and nine other zip codes had less foreclosures in October 2025 (can add the other zip codes too and locate those areas). There were many other zip codes that had zero change, specifically, some had 0 foreclosures in July 2021 and October 2025. For this question, I'll compare 20772 and 20618. It seems to be that 20772 is a zipcode in Prince George's County, covering most of Upper Marlboro. And 20618 is a zipcode in St. Mary's County, covering towns like Bushwood and Yellowbank. These areas are vastly differently since 20722 consists of a more suburban almost urban city and 20618 consists towns that are quite rural.

```{r}
# Put code here
foreclosure_change <- md_foreclosure_notices_zipcode |>
  select(zip, july_2021, october_2025) |>
  mutate(
    change = october_2025 - july_2021
  )

largest_increase <- foreclosure_change |>
  arrange(desc(change)) 

largest_decrease <- foreclosure_change |>
  arrange(change)

# Display results of code below this codeblock
largest_increase
largest_decrease
```

### Sentence 5

* **How many foreclosures are there in Maryland's lowest-income zip codes? How many foreclosures in high-income zip codes? Where are these areas located?**
* **XXXX**: In Maryland's ten highest median income zip codes, the number of notices of intent to foreclosure reported over the roughly 4-year time frame tracked by this dataset ranged from about 37 to 314 notices per 10,000 residents. The raw totals ranged from 4 to 558. These zip codes had a minimum household income ranging betwene 228,000 and 250,000 dollars per year. The zip codes are almost all suburbs in Montgomery and Howard counties, with the exception of 21056, which encompasses Gibson's Island outside Annapolis in Anne Arundel County. Comparatively, the 10 lowest median income zip codes saw averages ranging from zero to 653 notices served per 10,000 residents over the same time frame. The raw totals ranged from 0 to 1644 notices reported. Only one zip code had zero notices of intent to foreclose: 20890, which covers what appears to be a mostly-commercial area in Gaithersburg containing a single apartment complex. The zip with the next-lowest rate of foreclosures was 21835, which is located on the Eastern Shore and had a rate of about 130 per 10,000 residents. The low-income zip codes were mostly located in Baltimore, but there were also locations on the Eastern Shore and in Western Maryland. Overall, the lowest-income zip codes had a far higher number of notices of intent to foreclose than the high-income areas. Six out of the 10 lowest-median income zip codes had a rate of more than 400 notices reported per 10,000 residents. 

```{r}
# Put code here

county_income <- get_acs(
  geography = "zcta",
  variables = c(median_income = "B19013_001",
                total_pop = "B02001_001"),
  year = 2022,
  survey = "acs5",
  output="wide"
) |>
  rename(zip = GEOID) |>
  mutate(zip = as.character(zip)) 

foreclosures_by_income <- md_foreclosures_summary |> left_join(county_income, join_by(zip)) |> rename(total_foreclosures = total, median_income = median_incomeE, population = total_popE) |> mutate(foreclosure_rate=(total_foreclosures/population)*10000) 

foreclosures_lowest_income <- foreclosures_by_income |> select(zip, total_foreclosures, median_income, population, foreclosure_rate) |> arrange(median_income) 

foreclosures_highest_income <- foreclosures_by_income |> select(zip, total_foreclosures, median_income, population, foreclosure_rate) |> arrange(desc(median_income)) 

# Display results of code below this codeblock

foreclosures_highest_income 

foreclosures_lowest_income
```
***
Final Memo 

What’s the news? 

The pressure from rising foreclosure filings in Maryland is disproportionately affecting several predominantly Black communities. According to our analysis of state data, notices of intent to foreclose are sharply concentrated in a small number of zip codes, many of which are majority Black communities in Prince George’s County, where foreclosure notices have surged steadily since 2021. One Upper Marlboro-area zip code alone logged nearly 7,000 foreclosure notices in just over four years, far outpacing the rest of the state. 

So what? 

The pattern mirrors national reporting showing that foreclosure risk is rising fastest in communities that are already economically vulnerable. This data reveals exactly where that stress is landing in Maryland. 

This story is not about a statewide spike. It’s about geographic and racial concentration, and how some Maryland neighborhoods are facing significantly higher housing instability than others as pandemic-era protections fade. 


Preliminary analysis results

Our analysis of Maryland Notices of Intent to Foreclose filed in each zip code from July 2021 through October 2025 shows that foreclosure activity is highly concentrated in a few areas rather than broadly distributed. Zip codes in Prince George’s County, particularly 20772 and 20774, which are both located in and around Upper Marlboro, stand out dramatically from the rest of the state. Those areas have total notice counts several times higher than most other Maryland zip codes. They have also seen some of the largest increases over time, indicating that this is a long-term issue rather than short-term pressure.

When foreclosure data is joined with Census demographics, a clear racial pattern emerges. Nine of the ten zip codes with the highest foreclosure totals are majority Black, while zip codes with very low foreclosure activity are more likely to be majority White or have sparse residential populations. Income patterns reinforce this divide: The lowest-income zip codes experienced far higher foreclosure rates per capita than the highest-income zip codes.

These findings are consistent with widespread media reports on post-pandemic foreclosure risk, but they provide a localized, zip code-level view that has not been widely reported in Maryland. This data allows us to tell a story about where exactly in Maryland there is housing instability. And not just by the county or town, but potentially by specific neighborhoods as well.


Pros and cons of data
	
Pros: 
-The dataset covers every Maryland zip code and spans more than four years, allowing for both geographic and demographic comparisons (after using the census data), making it easy for trend analysis. 
-Monthly counts make it possible to identify sustained increases rather than isolated spikes. 
-By combining the foreclosures dataset with U.S. Census data, we can analyze racial, economic and other demographic disparities at the zip code or county level. 
-The data is published by the Maryland Office of Financial Regulation, lending it institutional credibility and consistency.

Cons: 
-The data reflects notices of intent to foreclose, not completed foreclosures or court filings, meaning it captures risk and pressure rather than outcomes.
-Zip codes vary widely in population and housing stock, so raw counts alone can be misleading without per-capita adjustments.
-No information is provided about property type, loan type or borrower circumstances. 
-The dataset could contain reporting errors and may not perfectly align with court foreclosure records. 
-Demographic and income data needed to be joined from external sources, which have margins of error and could require additional assumptions. 

Next steps to confirm findings

To confirm our analysis and further contextualize our findings, we would:
Compare notice counts with Maryland court foreclosure filings to assess how often notices lead to formal cases. 
Interview officials from the state's Office of Financial Regulation (which is under the Maryland Department of Labor) about the methodology used to collect the foreclosure notices, the quality of the data and any trends they are seeing. 
Talk with housing advocates, legal aid organizations and foreclosure prevention counselors in high-impact zip codes.
Request comment from county housing officials in Prince George’s County, Baltimore City and Baltimore County.
Seek response from lenders or banking associations if the data suggests concentrated lending risk. 
No claims of wrongdoing would be made without giving agencies and lenders a full and reasonable opportunity to respond. 

Reporting plan

To finish reporting the story, we would ground our analysis of the data in lived experiences. This would include interviewing homeowners and renters in the most affected zip codes, particularly in Prince George’s County, to understand how foreclosure pressure affects real households and the impact on local communities. We would speak with foreclosure counselors, housing attorneys and community organizations that work directly with residents facing foreclosure, such as from Community Legal Services of Prince George’s County’s Foreclosure Clinic. 

We would also visit at least one high-impact neighborhood, perhaps including one of the two in Upper Marlboro, to report on visible signs of housing distress, such as vacant properties or community stabilization efforts, and contrast that with conditions in high-income zip codes with minimal foreclosure activity. Joining zip code-level data with on-the-ground reporting and expert interviews would move the story beyond numbers and clearly show readers how unevenly housing instability is distributed across Maryland. 
***
