---
title: "Reverse Engineering Project"
author: Sam, Princess, Ellie
date: Oct. 18, 2025
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

In this notebook, we are reverse engineering the story, [Nearly 250 women have been
fatally shot by police since 2015] (https://www.washingtonpost.com/graphics/2020/investigations/police-shootings-women/)

## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}
library(tidyverse)
library(tidycensus)
library(janitor)
library(lubridate)
library(stringr)
```

## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary for data you are working with, and discuss any caveats or issues you discovered working with this data. 

**Answer** This data was compiled by The Washington Post using publicly-available government records, local news reports, social media and independent databases. After looking through the data, it appears very clean. The column names are all standardized, the date column is in the proper format, each entry is properly formatted for the column type. There don't appear to be duplicate entries. The only caveat is that this data is more recent that what The Washington Post used for its story. The article was published in 2020, but the database was continuously updated through December 2024. So even when filtering the data to only include values from before Sept. 4, 2020, they may have updated the dataset with more shootings that occurred during that time frame that were not included in their original analysis. Additionally, the story was published on Sept. 4, 2020, but it does not explicitly state when was the cutoff for the data included in their analysis. 


```{r}
# Load required data

# Path to data should be loaded from folder "data" i.e. read_csv("data/name_of_data.csv")

# Clean required data and prepare for analysis if needed. 

fatal_police_shooting_agencies <- read_csv("~/Documents/GitHub/data_journalism_2025_spring/major_assignments/reverse_engineering/data/fatal-police-shootings-agencies.csv")

fatal_police_shootings <- read_csv("data/fatal-police-shootings-data.csv") 
```

## Sentences to Engineer
In this notebook, we are reverse engineering five sentences from the story.

### Sentence 1

* **Sentence text**: The Post found only one other fatal shooting that closely matched Taylor’s case — a Black woman, unarmed, killed during a raid at home while a boyfriend shot at police... Of the 247 women fatally shot, 48 were Black and seven of those were unarmed.

* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

** We found that 243 women were fatally shot by the police between 2015-2020-09-04, according to data from the Post. 48 of them were Black and seven of them were unarmed. We can only confirm that 

```{r}
# Put code to reverse engineer sentence here
all_women <- fatal_police_shootings |>
filter(date <= as.Date("2020-09-04"), !is.na(race)) |>
select(race, gender, armed_with, name, date) |>
  filter(
    gender == "female"
  ) 

all_women <- all_women |>
    mutate(
    race = str_replace_all(race, "A", "asian"),
    race = str_replace_all(race, "B", "black"),
    race = str_replace_all(race, "H", "hispanic"),
    race = str_replace_all(race, "W", "white"),
    race = str_replace_all(race, "O", "other"), 
    race = str_replace_all(race, "N", "native american"),
  )



all_men <- fatal_police_shootings |>
filter(date <= as.Date("2020-09-04"), !is.na(race)) |>
select(race, gender, armed_with, name, date) |>
  filter(
    gender == "male"
  ) 
view(all_men)

woman_and_b <- fatal_police_shootings |>
filter(date <= as.Date("2020-09-04"), !is.na(race)) |>
  mutate(year = year(date)) |>
  filter(
    race == "B",
    gender == "female"
  ) |>
  select(race, gender, armed_with, name, date)

woman_and_b_unarmed <- fatal_police_shootings |>
filter(date <= as.Date("2020-09-04"), !is.na(race)) |>
select(race, gender, armed_with, name, date) |>
filter(str_detect(race, "B")) |>
filter(str_detect(gender, "female")) |>
  filter(str_detect(armed_with, "unarmed")) 

view(woman_and_b_unarmed)
# Display results of code below this codeblock
view(woman_and_b)

```

### Sentence 2

* **Sentence text**:  Black men, who are about 12 percent of the total U.S. male population, make up 27 percent of the men shot and 36 percent of the unarmed deaths. 

* **Analysis summary**: According to our analysis, Black men, who make up about 12 percent of the male population, make up roughly 26.4 percent of all men shot. Black men are more likely to be fatally shot by police. Among unarmed men shot, Black men make up more than 33.5 percent of those shot, meaning they are even more likely to be shot while unarmed. 

```{r}
# Put code to reverse engineer sentence here

us20 <- load_variables(2020, "acs5", cache=TRUE)

us_male_population <- get_acs(geography="us", variables=c(Black="B01001B_002", Total="B01001_002"), year=2020) 

black_male_population <- us_male_population |> select(variable, estimate) |> filter(variable=="Black") |> mutate(pct_total=(19716762/164545087)*100) 



black_men_total <- fatal_police_shootings |> filter(gender=="male" & !is.na(race) & date<="2020-09-04") |> group_by(race) |> summarise(total=n()) |> mutate(pct_shot_by_race=(total/sum(total)*100)) |>
  arrange(desc(pct_shot_by_race)) 

black_men_unarmed <- fatal_police_shootings |> filter(gender=="male" & armed_with=="unarmed" & !is.na(race) & date<="2020-09-04") |> group_by(race) |> summarise(total=n()) |> mutate(pct_shot_unarmed=(total/sum(total)*100))|>
  arrange(desc(pct_shot_unarmed))

# Display results of code below this codeblock

black_men_total

black_men_unarmed

black_male_population

```


### Sentence 3

* **Sentence text**: For both genders, a gun was the most common weapon. Of men killed, 57 percent were armed with a gun and of women, 44 percent. In other cases, women were armed with knives, cars, toy weapons, hammers and hatchets. By race, 147 of the women killed were White, 48 Black and 29 Hispanic. Five were Native American, four were Asian and three were other races. In 11 cases, race could not be determined. By percentages, men shot by police were less often White and more often Black or Hispanic. 

**Analysis summary**: Our analysis also shows that guns were the most common weapon that people fatally shot by police were armed with for both genders. 57.5 percent of males were armed with a gun, while 43.1 percent of females were armed with a gun. Other top weapons for females included knives, vehicles, and "replicas," which refer to toy weapons or replicas of weapons. By race, 150 of the women were White, 48 were Black, 32 were Hispanic, 6 were Asian, 5 were Native American, and 2 were listed as other. 

For men, while White males were the largest total group shop by police, they were less likely to be shot than Black males. For example, 26.4 percent of men fatally shot were Black, even though only about 12 percent of U.S. males are Black. Overall, our findings show similar trends to The Post's analysis,  but specific percentages and numbers vary even when filtered to the specific time period analyzed by The Post. 



```{r}
# Put code to reverse engineer sentence here
armed_with_male <- fatal_police_shootings |> filter(gender=="male"  & date <="2020-09-04") |> group_by(armed_with) |> summarise(incident_total=n()) |> arrange(desc(incident_total)) |> mutate(pct_male =(incident_total/sum(incident_total))*100)

armed_with_female <- fatal_police_shootings |> filter(gender=="female" & date <="2020-09-04") |> group_by(armed_with) |> summarise(incident_total=n()) |> arrange(desc(incident_total)) |> mutate(pct_female =(incident_total/sum(incident_total))*100)

females_by_race <- fatal_police_shootings |> filter(gender=="female" & !is.na(race) & date <="2020-09-04") |> group_by(race) |> summarise(incident_total=n()) |> arrange(desc(incident_total)) |>
  mutate(
    race = str_replace_all(race, "A", "asian"),
    race = str_replace_all(race, "B", "black"),
    race = str_replace_all(race, "H", "hispanic"),
    race = str_replace_all(race, "W", "white"),
    race = str_replace_all(race, "O", "other"), 
    race = str_replace_all(race, "N", "native american"),
  )

males_by_race <- fatal_police_shootings |> filter(gender=="male" & !is.na(race) & date <="2020-09-04") |> group_by(race) |> summarise(incident_total=n()) |> arrange(desc(incident_total)) |> mutate(race_pct =(incident_total/sum(incident_total))*100) |>
  mutate(
    race = str_replace_all(race, "A", "asian"),
    race = str_replace_all(race, "B", "black"),
    race = str_replace_all(race, "H", "hispanic"),
    race = str_replace_all(race, "W", "white"),
    race = str_replace_all(race, "O", "other"), 
    race = str_replace_all(race, "N", "native american"),
  )

black_male_population

# Display results of code below this codeblock
view(armed_with_male)

view(armed_with_female)

view(females_by_race)

view(males_by_race)
```


### Sentence 4

* **Sentence text**:
-As with the fatal shootings of men, the majority of women killed were armed.
-But women were more often fatally shot when unarmed [and]... 
-About 31 percent, or 77, of the 247 women fatally shot by police since 2015 had mental health issues, compared to 22 percent of the 5,362 men killed.

* **Analysis summary**: 
-- We found that 81.06 percent of women shot were armed (sum of gun, knife, vehicle, blunt_object and gun;knife) vs. 10.2 percent unarmed which aligns with the Post's finding. 
-- Our data shows that among people fatally shot when unarmed, 9.09% were women. And among people fatally shot while armed with a gun, 45.9% were women. 
-- We found that among all people fatally shot, women were more often fatally shot when unarmed at the time of the incident. And 6.8 percent of male victims were unarmed at the time of fatal police shootings compared to ten percent of women. 
-- We found that 82 or 33 percent of women shot, not 77 or 31 percent, had mental health issues at the time of shooting, compared to 23 percent of the 5,041 men killed. 

```{r}
# Put code to reverse engineer sentence here
armed_with_female <- fatal_police_shootings |>
filter(date <= as.Date("2020-09-04"), !is.na(race)) |>
filter(gender=="female") |> group_by(armed_with) |> 
summarise(incident_total=n()) |> arrange(desc(incident_total)) |> mutate(pct_female =(incident_total/sum(incident_total))*100)


total_gender <- fatal_police_shootings |>
filter(date <= as.Date("2020-09-04"), !is.na(race)) |>
  group_by(gender) |>
  summarise(total_gender_count = n()) 


unarmed_gender <- fatal_police_shootings |>
filter(date <= as.Date("2020-09-04"), !is.na(race)) |>
group_by( gender, armed_with) |>
filter(armed_with == "unarmed") |>
summarise(total_unarmed_gender_count = n()) 

unarmed_gender <- unarmed_gender |>
select(- armed_with) 



rates <- unarmed_gender |> left_join(total_gender, by="gender") |> mutate(rate_unarmed = total_unarmed_gender_count / total_gender_count * 100)

 

mental_related_case <- fatal_police_shootings |>
  filter(date <= as.Date("2020-09-04"), !is.na(race)) |>
  group_by(gender) |>
  summarise(
    total_cases = n(),  
    total_mental = sum(was_mental_illness_related == TRUE),  
    .groups = "drop"
  ) |>
  mutate(
    pct_mental = total_mental / total_cases * 100)



# Display results of code below this codeblock
view(armed_with_female)
view(total_gender)
view(unarmed_gender)
view(rates)
view(mental_related_case)
```


### Sentence 5

* **Sentence text**:  Since 2015, Black women have accounted for less than 1 percent of the overall fatal shootings in cases where race was known. But within this small subset, Black women, who are 13 percent of the female population, account for 20 percent of the women shot and killed and 28 percent of the unarmed deaths.

* **Analysis summary**: 
--We found that Black women accounted for 0.9% of the overall fatal shooting cases where race was known, which is aligned with the Post's findings of less than 1%. 
-- Our findings show Black women make up 12.7% of the total population, closely aligned with the Post's finding. They also accounted for 19.2 percent of the women shot and killed. 
--We found Black women made up 1.8 percent, 7 total, of the unarmed deaths, not aligned with the Post's finding of 28 percent. We are not sure what happened here as women only make up 6.7 percent of unarmed deaths, per our analysis. So we can't say why the Post said Black women made up 28 percent of unarmed deaths when they didn't. 
--Our findings instead showed that Native, Asian, and Black men had the highest percentages for unarmed deaths. 

Count incidents by combination of race + gender 
```{r}
# Put code to reverse engineer sentence here

#overall percentage of shootings by race and gender
overall_pct <- fatal_police_shootings |>
filter(date <= as.Date("2020-09-04"), !is.na(race)) |>
  group_by(race, gender) |>
  summarise(count = n(), .groups = "drop") |>
  mutate(percent = (count / sum(count)) * 100)

overall_pct <- overall_pct |>
  mutate(
    race = str_replace_all(race, "A", "asian"),
    race = str_replace_all(race, "B", "black"),
    race = str_replace_all(race, "H", "hispanic"),
    race = str_replace_all(race, "W", "white"),
    race = str_replace_all(race, "O", "other"), 
    race = str_replace_all(race, "N", "native american"),
  )

#Percentage of women fatally shot by race
women_shot_pct <- fatal_police_shootings |>
filter(date <= as.Date("2020-09-04"), !is.na(race)) |>
  filter(gender == "female") |>
  group_by(race) |>
  summarise(count = n(), .groups = "drop") |>
  mutate(percent = (count / sum(count)) * 100)

women_shot_pct <- women_shot_pct |>
   mutate(
    race = str_replace_all(race, "A", "asian"),
    race = str_replace_all(race, "B", "black"),
    race = str_replace_all(race, "H", "hispanic"),
    race = str_replace_all(race, "W", "white"),
    race = str_replace_all(race, "O", "other"), 
    race = str_replace_all(race, "N", "native american"),
  )


#percent of total unarmed shootings by race and gender
unarmed_pct <- fatal_police_shootings |>
  filter(date <= as.Date("2020-09-04"), !is.na(race)) |>
  filter(armed_with == "unarmed") |>
  group_by(race, gender) |>
  summarise(count = n(), .groups = "drop") |> 
  mutate(percent = count / sum(count) * 100)

unarmed_pct <- unarmed_pct |>
   mutate(
    race = str_replace_all(race, "A", "asian"),
    race = str_replace_all(race, "B", "black"),
    race = str_replace_all(race, "H", "hispanic"),
    race = str_replace_all(race, "W", "white"),
    race = str_replace_all(race, "O", "other"), 
    race = str_replace_all(race, "N", "native american"),
  )

view(unarmed_pct)

#percent of total unarmed shootings by gender
unarmed_pct_gender <- fatal_police_shootings |>
  filter(date <= as.Date("2020-09-04"), !is.na(race)) |>
  filter(armed_with == "unarmed") |>
  group_by(gender) |>
  summarise(count = n(), .groups = "drop") |> 
  mutate(percent = count / sum(count) * 100)

view(unarmed_pct_gender)

# Black women vs total population
us23 <- load_variables(2023, "acs5", cache=TRUE)

us_female_population <- get_acs(geography="us", variables=c(Black="B01001B_017", Total="B01001_026"), year=2023) 
  
black_female_population <- us_female_population |> 
select(variable, estimate) |> 
tidyr::pivot_wider(
    names_from = variable,
    values_from = estimate
  ) |>
  mutate(pct_total=(Black / Total)*100)

black_female_population
# Display results of code below this codeblock

view(overall_pct)
view(women_shot_pct)
view(unarmed_pct)
view(black_female_population)
```

-30-