---
title: "lab_03"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## TASK: Load libraries

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse
library(tidyverse)

# Load janitor 
install.packages('janitor')
library(janitor)
```

## TASK: Load data and clean up the column names

Load the two datasets we worked with in the pre_lab that are in the data folder: the Maryland payments to vendors and Maryland grants and loans AND, using janitor, change the column names so that they are lower-cased and any spaces are replaced with underscores.

```{r}
# Load in the Maryland state agency payments data, making sure that it correctly sets the datatypes using all of the rows
payments <- read_csv("~/Documents/GitHub/data_journalism_2025_spring/labs/lab_03/data/State_of_Maryland_Payments_Data__FY2008_to_FY2024.csv", guess_max=369008)

# Load in the Maryland grants and loan data
md_grants_loans <- read_csv("~/Documents/GitHub/data_journalism_2025_spring/labs/lab_03/data/State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2022.csv")

payments <- payments |> clean_names()

md_grants_loans <- md_grants_loans |> clean_names()

```

## Answer questions

Q1: You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed in the payments data. Look at & search the records in the dataframe to start your search, starting with a broad search and then narrowing it down. Then write code to find rows where the vendor name contains some version of the county's school system.

HINT: you should use filter() just once, and inside your filter you want to look for the variations one at a time. Think "this version OR that version."

When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find?

A1: I found 92 records for the school system. 

```{r}
pg_schools <- payments |> filter(vendor_name == "PRINCE GEORGES COUNTY PUBLIC SCHOOLS" | vendor_name == "PRINCE GEORGE'S COUNTY PUBLIC SCHOOLS" | vendor_name == "PRINCE GEORGES COUNTY PUBLIC SCHOOL" |vendor_name == "PG COUNTY PUBLIC SCHOOL" | vendor_name == "PRINCE GEORGES PUBLIC SCHOOLS" | vendor_name == "PRINCE GEORGES COUNTY SCHOOL"| vendor_name == "BOARD OF EDUCATION OF PRINCE GEORGE" | vendor_name == "BOARD OF EDUCATION OF PRINCE GEORGE'S COUNTY" | vendor_name == "BOARD OF EDUCATION OF PRINCE GEORGES COUNTY MD")
```

Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? Are there any aspects of the results that stand out to you?

A2: Fiscal year 2024 had the largest total amount. The main aspect that stands out to me is that the total amount had a significant increase each fiscal year. 

```{r}
pg_schools |> 
  group_by(fiscal_year) |>
  summarise(
    total_amount = sum(amount)
  ) |>
  arrange(desc(total_amount))
```

Q3. You're interested in examining the vendors paid for capital projects at the University of Maryland. Search the payments data to find those payments, then write code to produce a list of each vendor and the total amount it received.

Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? Provide at least three examples of vendor names that are similar and should be combined into one. Describe in words how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?

A3. I would use the distinct() function to eliminate duplicate values of specific rows of data. There are several vendors listed at the same zip but with slight alterations to their name in the data. "VWR INTERNATIONAL LLC" is listed at the same zip as "VWR INTERNATIONAL INC", "GE HEALTHCARE BIO-SCIENCES CORP" is listed at the same one as "GE HEALTHCARE BIO-SCIENCES CORP," and "WDG ARCHITECTURE PLLC" is listed alongside "WDG ARCHITECTURE." These are all values that could potentially be combined. 

```{r}
UMD_capital_projects <- payments |> filter(agency_name =="CAPITAL PROJECTS- UNIVERSITY OF MARYLAND") |>
  group_by(vendor_name, vendor_zip) |>
  summarise(
    total_amount = sum(amount)
  ) |>
  arrange(desc(vendor_name))
```

Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year. Search the data to find grants that fit that description. Once you find which column(s) contain the information you want, your code should find just the grants for stem cell research in a single filter. Then calculate the total amount of money for each fiscal year. Write a sentence summarizing how stem cell research funding has changed over time.

A4: Funding for stem cell research in Maryland mostly varied between about 4.8-9.2 million dollars each year, but had peaks of 14.6 million and 12.4 million dollars in 2009 and 2023, respectively. 

```{r}
stem_cell_grants <- md_grants_loans |> mutate(description = str_to_upper(description)) |> filter(str_detect(description, "STEM CELL")) |>
  group_by(fiscal_year) |>
  summarise (
    total_amount = sum(amount)
  ) |>
  arrange(desc(fiscal_year))
```

Q5: Using the Maryland grants and loan data, find records where the zip code is the one where the University of Maryland-College Park is located in and produce a list of grantees from that zip code, along with the total amount they received. Write a paragraph explaining how many grantees are there, what you would need to do to clean the results so that you had one record per grantee and which grantees are most interesting to you?

A5: There are 50 total grantees that are listed at the University of Maryland, College Park's sip code. I would use the distinct() function to save only value for each specific record. I found the Branchville Fire Dept., the Pregnancy Aid Center and Ikea as the most interesting grantees listed in the data. 

```{r}
UMCP_zip <- md_grants_loans |> filter(zip_code == "20740" | zip_code == "20742") |>
  group_by(grantee, zip_code) |>
  summarise(
    total_amount = sum(amount)
  ) |>
  arrange(desc(total_amount))
  
```
