---
title: "lab_05"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   Tabula

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse, plus any other packages you will need to clean data and work with dates.
library(tidyverse)
library(janitor)
library(lubridate)

```

## Get Our PDF

We'll be working with the [911 overdose calls from Baltimore County](https://drive.google.com/file/d/1qkYuojGF_6WKFr5aNQxmewDzcKyOiJFr/view?usp=share_link). You'll want to download it to a place you'll remember (like your Downloads folder, or the labs folder in your repository). The goal is to extract the tables within it, export that to a CSV file, load it into RStudio and ask some questions.

## Extract Data from PDF Using Tabula

Start Tabula, then go to <http://127.0.0.1:8080/> in your browser. Click the "Browse" button and find the PDF file and click "open", and then click the "Import button" in Tabula. This will take a few seconds or longer.

This PDF has a single table spread over multiple pages to extract. We're going to make a single dataframe from this table, exporting it to a CSV file that you will load into R. In Tabula, highlight the table and click the "Preview & Export Extracted Data" button. You may want to play with including or excluding the column headers - YOU SHOULD HAVE FIVE COLUMNS OF DATA.

Save the CSV (it should be called `tabula-Baltimore County; Carey, Samantha log OD.csv` by default) to your lab_05/data folder.

From there, you will need to read in the data, and add or fix headers if necessary. You can choose to include the headers from the PDF in your exported CSV files OR to exclude them and add them when importing. `read_csv` allows us to do this ([and more](https://readr.tidyverse.org/reference/read_delim.html)).

## Load and clean up the data in R

You will need to read in and clean up the data so that it can be used for analysis. By "clean" I mean the column headers should not contain spaces and they should have meaningful names, not "x1" or something similar. How you do that is up to you, but you can use select() with or without the minus sign to include or exclude certain columns. You also can use the `rename` function to, well, rename columns. Importantly, you'll need to ensure that any columns containing a date actually have a date datatype. Our friend `lubridate` can help with this.

```{r}
Baltimore_County_overdoses <- read_csv("~/Documents/GitHub/data_journalism_2025_spring/labs/lab_05/tabula-Baltimore County; Carey, Samantha log OD.csv", col_names = FALSE) |> clean_names() |>
  rename(DAT_FORMAT = x1, DAT_FORI = x2, CASE_NBR = x3, EVTYP = x4, LOC = x5) |> mutate(DAT_FORMAT=mdy(DAT_FORMAT))
```

## Answer questions

Q1. Write code to generate the number of calls that occurred on each date. Which date in 2022 had the most overdose calls, and how many? Look at the total number of rows in your result and explore the range of dates - based on your result, do you believe there are any days with no overdose calls at all? Explain why or why not.

A1. July 14 and Oct. 4 are tied for the most overdoses in 2022 at 23. I don't believe there are any days without overdoses. The data spans from Feb. 6, 2022 to Feb. 6, 2023. When grouping the dataset by day, there are 366 rows, meaning there is an entry for each day in the timespan. 

```{r}
data_by_date <- Baltimore_County_overdoses |> 
  group_by(DAT_FORMAT) |>
  summarize(total_calls = n()) |>
  arrange(desc(total_calls))

```

Q2. You want to understand if there's a pattern in the day of the week that overdose calls are made. Add a column to your dataframe that displays what day of the week each date represents. You should search for how to do that using lubridate. Then write code to calculate the number of calls for each day of the week, and add a column to that result that calculates the percentage of all calls that occurred on each day of the week (so you want a dataframe with the day of the week, total number of calls and the percentage of calls on that day out of the total number of all calls). Describe your findings to me.

A2. Saturday was the most frequent day for overdoses, with 638 calls totaling about 15.53% of all overdoses. Overall, the weekend was the most common time for overdoses, with Saturday, Sunday and Friday being the top 3 days. Thursday is the lowest of the week, with 522 calls or about 12.7% of the total. 

```{r}
Baltimore_County_overdoses <- Baltimore_County_overdoses |> mutate(day_of_week = wday(DAT_FORMAT, label = TRUE, abbr = FALSE)) 
data_by_DOTW <- Baltimore_County_overdoses |>
  group_by(day_of_week) |>
  summarize(day_frequency = n()) |>
  arrange(desc(day_frequency)) |> 
  mutate(percent_of_calls = 100 * day_frequency / sum(day_frequency))
```

Q3. Now let's look at locations. Which ones have the most calls? How would you describe them (feel free to search for more information on them)? Is there anything about the structure of the original data that might make you less confident in the counts by location or date?

A3. The top five locations with the most calls are 4540 SILVER SPRING RD, PC 02; 6424 WINDSOR MILL RD, PC 06; 115 SUSQUEHANNA AV W, 330 LEEANNE RD, and 5018 PAINTERS MILL RD. Of those locations, two are Police precincts and one is the Owings Mills train station. Of the other two locations, according to Google Earth and Zillow, 4540 SILVER SPRING RD seems to be a single-family residence outside Whitemarsh and 330 LEEANNE RD is a rowhouse outside Essex. My only thought about the locations is that the listed addresses only list the street address, they don't include the zip code or state. Regarding the time of the calls, there are several listed as coming in at exactly 00:00:00 which makes me wonder if there was no time entered for those calls and that is the default selected time. 

```{r}
data_by_loc <- Baltimore_County_overdoses |> 
  group_by(LOC) |>
  summarize(location_frequency = n()) |>
  arrange(desc(location_frequency))
```

Q4. What's the best story idea or question you've seen as a result of the work you've done in this lab?

A4. I think the best story idea I found is that there was a reported overdose every single day during the span of the data, and only two days had a single call. After looking through the data of overdose calls by date, I noticed that there are spans where there will be around 15 calls or so per day for several days or about a week, and there are spans where there are about 5 or so calls per day for several days or a week. I'm curious what is driving that fluctuation. 
