---
title: "lab_10"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   Our usual libraries, plus some of the map-specific ones from the pre_lab. You'll also be fetching Census data in Q3. Read the questions carefully!

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}
# Load libraries
library(tidyverse)
library(tidycensus)
library(sf)
library(janitor)
library(tigris)

```

Let's explore Maryland elections, power outages and electric vehicle data and make some maps!

## Questions

**Q1.** Maryland voters made history last November by electing the state's first two African-American statewide office holders: Wes Moore as governor and Anthony Brown as attorney general. You want to compare the two's electoral performances. Using the `md_statewide_county.csv` file in the data folder, calculate the total number of votes for both the governor's race and the attorney general's race, then percentages for both Moore and Brown *based on the total of their respective contests*. Then calculate the difference between their percentages.

Get the geometry for Maryland's counties using tigris and, joining that dataframe to your results dataframe, make a county-level map of the differences between Moore and Brown's percentages. What is the story here, and what is the county with the most interesting results?

**A1.**The story here is that Moore generally out performed Brown across the state. Brown only received a larger share of the vote in Prince George's County and Batlimore City. Moore especially did better in more conservative/rural areas. Moore saw his largest difference in the share of the vote in Queen Anne's County, where he about 7.4 percent more of the vote than Brown. Comparatively, Brown outperformed Moore by 1.6 percent in Prince George's by about 1 percent in Baltimore. I think those are the most interesting counties as they are two of the state's largest and the only ones in which Brown did better. 

```{r}
election_results <- read_csv("~/Documents/GitHub/data_journalism_2025_spring/labs/lab_10/data/md_statewide_county.csv") |> clean_names()

ag_gov <- election_results |> group_by(geoid) |>
  summarise (
    total_gov = sum(cox, moore, lashar, wallace, harding, sefcik, gov_writein),
    total_ag = sum(peroutka, brown, ag_writein),
    pct_moore = (moore/total_gov)*100,
    pct_brown = (brown/total_ag)*100,
    pct_vote_diff = pct_moore - pct_brown,
    geoid = as.character(geoid)
  ) 


md_geom <- counties() |> filter(STATEFP == "24") |> clean_names()  |> rename(county=namelsad) |> mutate(county = str_remove(county, " County")) |> mutate(county = str_to_title(county)) 


ag_gov_mapped <- md_geom |> left_join(ag_gov, join_by(geoid)) |> select(geoid, county, total_gov, total_ag, pct_moore, pct_brown, pct_vote_diff, geometry) 
  

ggplot() +
  geom_sf(data=ag_gov_mapped, aes(fill = pct_vote_diff)) +
  scale_fill_gradient2(
    midpoint = 0
  ) +
  theme_minimal() +
  labs(
    title = "Wes Moore generally outperformed Anthony Brown in 2022",
    subtitle = "Brown received a larger share of the vote in only two counties",
    fill = "Moore's vote difference (%)"
  ) 

```


**Q2.** Make a county-level map to explore the number of customers impacted by power outages in Maryland from last March. Load the `county_outages_march2023.csv` file in the data folder and calculate a rate of customers with outages per 10,000 total customers for each county. Then join that dataframe to your Maryland counties dataframe you made in Q1 and create a county-level map of the results; you may need to tweak the fill scale to make the distinctions clearer. What's the potential story here? Given access to other months' outage data, what would your next steps be in reporting a story?

**A2.**A potential story here is Talbot county seeing a far higher outage rate compared to any other Maryland county. Talbot had about 1990 customers per 10,000 affected by outages. The next highest county, Charles, only had about 257. Given other months data, I would see if Talbot was the still the highest then, too, if is this month was some outlier. If it was still the highest, then that would definitely be a story. It'd be interesting to note changes over time as well, and see if this issue is getting better, worse, staying the same, etc. 

```{r}
county_outages <- read_csv("~/Documents/GitHub/data_journalism_2025_spring/labs/lab_10/data/county_outages_march2023.csv") |> clean_names() 


county_outages <- county_outages |> group_by(county) |> 
  summarise (
    outages_rate = (outage_customers/total_customers)*10000) 


outages_mapped <- md_geom |> left_join(county_outages, join_by(county)) |> select(county, outages_rate, geometry) 


ggplot() +
  geom_sf(data=outages_mapped, aes(fill = outages_rate)) +
  scale_fill_viridis_b(option="magma", trans = "log") +
  theme_minimal() +
  labs(
    title = "Nearly 2 out of 10 Talbot County customers impacted by power outages",
    subtitle = "Maryland's 23 other counties ranged between 1 and 257 customers impacted per 10,000",
    fill = "Outages per 10,000 customers") 

```

**Q3.** You've been assigned a story about the use of electric vehicles in Maryland and you want to make a map of the prevalence of electric vehicles in each county. Load the `ev_2020_2023.csv` file, which has the number of electric and hybrid vehicles per county from July 2020 and March 2023 and calculate a percentage change for each county (remember, that's `new-old/old`). Using the Maryland counties dataframe you made for Q1, make a county map showing the percentage change. What counties stand out to you? What questions do you have about the changes you're seeing?

Next, use tidycensus to retrieve county population data for Maryland from 2021 (you did this in the pre_lab) and join it to your vehicles dataframe. HINT: you may not need to join it on the county name, and you should already have the geometry!

Then, calculate a per capita rate of electric & hybrid vehicles using the March 2023 figure and the population estimate. You will need to choose the "per capita" value.

Finally, make a second county map showing the per capita rate. What counties stand out on this map? How is it different from the first one you made?

**A3.** Montgomery and Howard stand out as they had the highest per capita increase. That's different than the first result showing percent change, which had Caroline county as the highest percent increase at about 426 percent. 

```{r}
md_ev_hybrids <- read_csv("~/Documents/GitHub/data_journalism_2025_spring/labs/lab_10/data/ev_2020_2023.csv") |> clean_names() 


ev_hybrids_change <- md_geom |> left_join(md_ev_hybrids, join_by(county)) |> group_by(county) |>
  mutate(
    pct_change = ((march_2023-july_2020)/july_2020)*100
  ) |> select(county, july_2020, march_2023, pct_change, geometry) 


md_county_population <- get_acs(geography = "county",
              variables = c(population = "B01001_001"),
              year = 2021,
              state = "MD") |> clean_names() |> rename(county=name) |>
  mutate(county = str_remove(county, " County, Maryland"),
         county = str_remove(county, ", Maryland")) |> mutate(county = str_to_title(county)) 


ev_hybrids_change_mapped <- ev_hybrids_change |> left_join(md_county_population, join_by(county)) |> group_by(county) |> mutate(increase_per_capita = (march_2023/estimate)*1000) |>
  select(county, july_2020, march_2023, pct_change, increase_per_capita, estimate, geometry) 


ggplot() +
  geom_sf(data=ev_hybrids_change_mapped, aes(fill = increase_per_capita)) +
  scale_fill_viridis_b(option="magma", trans = "log") +
  theme_minimal() +
  labs(
    title = "Howard County sees largest increase in EV and Hybrid vehicle use",
    subtitle = "The number of EVs and Hyrbrids increased in each county from July 2020 to March 2023",
    fill = "Increase in vehicles per 1000 residents") 
  
```
